How many speech sounds are there in the world's languages?

### 2023-08-04 README.Rmd
updated SI
updates
submitted version

---
title: "Supplementary materials for: How many speech sounds are there in the world's languages?"
author: "Steven Moran & Nicholas A. Lester\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
      toc: true
bibliography: 'bibliography.bib'
---

@@

```{r}
plot(segments_upsid$Rank, segments_upsid$Frequency, xlab = "Rank", ylab = "Frequency", main = "UPSID-451 segment distribution")
plot(segments_phoible$Rank, segments_phoible$Frequency, xlab = "Rank", ylab = "Frequency", main = "PHOIBLE phoneme distribution")
```

@@

# Estimating the parameters -- TODO have set c lower
fitted <- nls(y ~ zipf(x_rank, a, b, c), data = df, start = list(a = -1.3, b = 18, c = 100000), trace = T)

@@

# Settings of K and B based on trial and error -- not the optimal values, but the result is reasonably close to the true value. Therefore, if we do any model tuning, the values we arrive at should be close to these; otherwise, we most likely have a bad fit of Herdan-Heaps' to the data. 
heaps(8, n, .5)
```

@@

@@

# Function to estimate the number of unseen types from a proposed assemblage (used to compute our estimate of Q0)
S_Chao2 <- function(vector, R) {
  fd <- vector[vector > 0]
  Q1 <- sum(vector == 1)
  Q2 <- sum(vector == 2)
  Sobs <- length(fd)
  if (Q2 > 0.0) {
    res <- Sobs + ((R - 1.0) / R) * ((Q1**2.0) / (2.0 * Q2))
  }
  else {
    res <- Sobs + ((R - 1.0) / R) * Q1 * (Q1 - 1) / 2.0
  }
  return(res)
}
```

@@

# r = 2000
chao.iter.plot.2000 <- choa.iter.plotter(chao.data.2000.10, mean.obs.df.2000, "r = 2000") + ylim(0,4000)
chao.iter.plot.2000
```

>>>>

---
title: "Supplementary materials for: How many speech sounds are there in the world's languages?"
author: "Steven Moran & Nicholas A. Lester\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: true
    latex_engine: xelatex    
bibliography: 'bibliography.bib'
biblio-style: "apalike"
csl: unified-style-sheet-for-linguistics.csl
link-citations: true
---

@@

```{r}
plot(segments_upsid$Rank, segments_upsid$Frequency, 
     xlab = "Rank", 
     ylab = "Frequency", 
     main = "UPSID-451 segment distribution")

plot(segments_phoible$Rank, segments_phoible$Frequency, 
     xlab = "Rank", 
     ylab = "Frequency", 
     main = "PHOIBLE phoneme distribution")
```

@@

# Estimating the parameters
fitted <- nls(y ~ zipf(x_rank, a, b, c), 
              data = df, 
              start = list(a = -1.3, b = 18, c = 100000), 
              trace = T)

@@

# Settings of K and B based on trial and error -- not the optimal values, but 
# the result is reasonably close to the true value. Therefore, if we do any 
# model tuning, the values we arrive at should be close to these; otherwise, 
# we most likely have a bad fit of Herdan-Heaps' to the data. 
heaps(8, n, .5)
```

@@

\newpage

@@

# Function to estimate the number of unseen types from a proposed assemblage 
# (used to compute our estimate of Q0)
S_Chao2 <- function(vector, R) {
  fd <- vector[vector > 0]
  Q1 <- sum(vector == 1)
  Q2 <- sum(vector == 2)
  Sobs <- length(fd)
  if (Q2 > 0.0) {
    res <- Sobs + ((R - 1.0) / R) * ((Q1**2.0) / (2.0 * Q2))
  }
  else {
    res <- Sobs + ((R - 1.0) / R) * Q1 * (Q1 - 1) / 2.0
  }
  return(res)
}
```

@@

# r = 2000
chao.iter.plot.2000 <- 
  choa.iter.plotter(chao.data.2000.10, mean.obs.df.2000, "r = 2000") + 
  ylim(0,4000)

chao.iter.plot.2000
```

### 2023-08-04 paper.Rmd
submitted version

---
title: "How many speech sounds are there in the world's languages?"
author: "Steven Moran & Nicholas A. Lester\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    number_sections: true
bibliography: 'bibliography.bib'
biblio-style: "apalike"
csl: unified-style-sheet-for-linguistics.csl
link-citations: true
header-includes:
  \usepackage{fontspec}
  \setmainfont{Times New Roman}
  \setmonofont{Arial Unicode MS}
  \usepackage{lineno}
  \linenumbers
---

@@

How many speech sounds are there in the world's languages? To answer this question, we extend two formal models of lexical diversity to segmental phonology. Zipf's law states that most word types are rare; Herdan-Heaps' law states that vocabulary size grows indefinitely with increasingly new text. We find that both models produce accurate estimates of the diversity of contrastive speech sounds (phonemes) in existing typological databases. Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) documenting more phonological repertoires cross-linguistically always yields more phoneme types. As a sanity check, we apply an unrelated estimator of species diversity from biology, Chao-Shen, and show that its estimates align with Herdan-Heaps' law. We use both models to extrapolate phoneme counts to undocumented languages, resulting in an estimate of 4600-4900 distinct phonemes in the full set of languages thought to be spoken today (~7-8k), which is roughly 1.5x more than what is currently documented in the world's languages (~35%). By extrapolating to hypothetical sample sizes, e.g., the number of languages estimated to have been spoken through time, and beyond, we find no indication of convergence: infinite languages mean infinite phonemes.

@@

Assuming a target of 7151 known languages,^[https://www.ethnologue.com/guides/how-many-languages?] our models predict a range of total segment types from 4592 to 4850 (mean = 4721). Based on PHOIBLE, our current best guess of the diversity of documented phoneme types is 3183. Therefore, the models argue for the existence of approximately 1500 yet to be documented phoneme types. As we probe increasingly larger and theoretically motivated language sample sizes, i.e., languages that existed *through time* (e.g., estimates including 30k-500k [@Pagel2000] or 64k-140k [@Crystal2002]), the Herdan-Heaps' model predicted a continual increase in the number of unique phonemes, reaching values as high as 19K. The Chao-Shen model produced much more conservative estimates for samples of size 10K and above, even appearing to settle on an asymptote of roughly 5500 phonemes for samples of 100K languages or more. However, when we compare the predictions of Herdan-Heaps and Chao-Shen  against known type counts in a closed dataset, the source of the discrepancy emerged: as the sample grows, Chao-Shen lags behind Herdan-Heaps in approximating the true value. Thus, given a larger pool of training data, we expect Chao-Shen to approximate more closely the estimates of Herdan-Heaps' law.

We therefore provisionally conclude three things: (1) adding languages to a sample will always carry a non-zero probability of producing a novel phoneme; (2) in accordance with Herdan-Heaps' law, the rate at which new phonemes are discovered per language will always decrease; and (3) the decrease in the rate of new observations will never reach 0. Thus, based on the evidence presented here, we have no reason to believe that there is an upper bound on the number of unique phonemes possible in spoken languages. Infinite linguistic diversity seems to breed infinite possible phonemes. But what would it mean if there were indeed infinitely many potential phonemes?

If one adopts a feature-based perspective on the phonetic targets of phonemes, then there is an *a priori* upper bound of $2^F$ contrastive sounds, where $F$ is the number of feature categories and 2 reflects the choice between presence (+) or absence (-) of that feature.^[In some feature sets, some values may be 0, i.e., undefined, underspecified, or ``for which a segment can be said 'not to care'" [@Hayes2009, pg. 91]. For example, speech sounds that are -coronal are underspecified for the features +/-anterior and +/-distributed, which capture coronal distinctions for places of articulation involving (inter-)dentals and alveolars vs palatal alveolars and retroflexes (anterior), and laminal vs apical coronals (distributed).] @Clements2009 coined this calculation *Feature Bounding*. The expected number of unique phonemes is controlled by the number of features, where additional features increase the pool of possible distinctive segments (nearly) exponentially. For example, given a feature set such as in PHOIBLE with 37 distinctive features, an upper bound on the number of segments under a strictly binary formalization is over 137 billion (2^37) feature vectors.

@@

However, most widely accepted feature sets contain features that are specified at different levels of phonetic detail: some features are more abstract (i.e., general similarities such as +/- syllabic), while others are more concrete (i.e., closer to the articulatory reality, e.g., advanced tongue root). Some are grounded in pure articulatory or acoustic phonetics, while others involve phonological properties of language. And looming in the background of these discussions, the debate of whether distinctive features are innate and universal or rather learned and emergent continues. Nevertheless, in general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors (2^24). Current consensus aside, there does not appear to be a principled limit to the degree of phonetic specificity that can be encoded by a feature. Making finer and finer contrasts may always produce more and more features. Thus, the exponent in the Feature Bounding calculation can itself (by hypothesis) be infinite, in line with the predictions of the models presented here, and therefore with the notion of infinite possible phonemes. That said, contemporary feature sets can capture phonological classes for place and manner of articulation, which neatly classify the phonological patterns and processes in and across languages (cf. @Mielke2008).

@@

We have so far seen that the feature space is a discrete re-coding of the continuous variability of the physical acoustic and articulatory spaces -- and so it should by way of finer and finer specification, be able to support an infinitely diverse set of phonemes. However, one objection to this account is that humans are subject to physiological limits, both in production and perception. In speech production, articulatory gestures are executed with some degree of imprecision [@Ladefoged_etal1972;@Lindblom1990]. If this stochastic variability is greater than the difference between two or more phonetic targets, then no reliable distinction can be sustained (cf. @Stevens1972). In comparison, if the range of articulatory gestures is greater than their acoustic (or perceptual) reception, there is a "many-to-one" mapping between articulation and perception. In perception, the ability to discriminate acoustic stimuli is also limited both anatomically and cognitively. Differences in the signal that fall below the various thresholds for discrimination are thus ignored.

@@

Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond whatever the phonetics-phonology interface may or may not be. Thus, we believe the approach presented here has broad applications to the study of other areas of grammar and to the cross-linguistic study of languages.

>>>>

---
title: "How many speech sounds are there in the world's languages?"
author: |
  | Steven Moran^1,2^ & Nicholas A. Lester^3^
  | ^1^Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
  | ^2^Department of Anthropology, University of Miami, Coral Gables, FL, USA
  | ^3^Department of Linguistics, University of North Texas, Denton, TX, USA
  | OrcID, SM, 0000-0002-3969-6549
  | Corresponding authors: steven.moran@unine.ch; nicholas.a.lester@gmail.com
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    number_sections: true
bibliography: 'bibliography.bib'
biblio-style: "apalike"
csl: unified-style-sheet-for-linguistics.csl
link-citations: true
header-includes:
  \usepackage{fontspec}
  \setmainfont{Times New Roman}
  \setmonofont{Arial Unicode MS}
  \usepackage{lineno}
  \linenumbers
---

@@

How many speech sounds are there in the world's languages? To answer this question, we extend two formal models of lexical diversity to segmental phonology. Zipf's law states that most word types are rare; Herdan-Heaps' law states that vocabulary size grows indefinitely with increasingly new text. We find that both models produce accurate estimates of the diversity of contrastive speech sounds (phonemes) in existing typological databases. Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) increased documentation of phonological inventories always yields more phoneme types. To check this, we apply an unrelated estimator of species diversity from biology, Chao-Shen, and show that its estimates also align with Herdan-Heaps' law. We use both models to extrapolate phoneme counts to undocumented languages, resulting in an estimate of 4600-4900 distinct phonemes in the full set of languages thought to be spoken today (~7-8k), which is roughly 1.5 time more than what is currently documented in the world's languages (~35%). By extrapolating to hypothetical sample sizes, e.g., the number of languages estimated to have been spoken through time, and beyond, we find no indication of convergence: infinite languages mean infinite phonemes.

@@

Assuming a target of 7151 known languages,^[https://www.ethnologue.com/guides/how-many-languages?] our models predict a range of total segment types from 4592 to 4850 (mean = 4721). Based on PHOIBLE, our current best guess of the diversity of documented phoneme types is 3183. Therefore, the models argue for the existence of approximately 1500 yet to be documented phoneme types. As we probe increasingly larger and theoretically motivated language sample sizes, i.e., languages that existed *through time* (e.g., estimates including 30-500k [@Pagel2000] or 64-140k [@Crystal2002]), the Herdan-Heaps' model predicted a continual increase in the number of unique phonemes, reaching values as high as 19k. The Chao-Shen model produced much more conservative estimates for samples of size 10k and above, even appearing to settle on an asymptote of roughly 5500 phonemes for samples of 100k languages or more. However, when we compare the predictions of Herdan-Heaps and Chao-Shen  against known type counts in a closed dataset, the source of the discrepancy emerged: as the sample grows, Chao-Shen lags behind Herdan-Heaps in approximating the true value. Thus, given a larger pool of training data, we expect Chao-Shen to approximate more closely the estimates of Herdan-Heaps' law.

We therefore provisionally conclude three things: (1) adding languages to a sample will always carry a non-zero probability of producing a novel phoneme; (2) in accordance with Herdan-Heaps' law, the rate at which new phonemes are discovered per language will always decrease; and (3) the decrease in the rate of new observations will never reach 0. Thus, based on the evidence presented here, we have no reason to believe that there is an upper bound on the number of unique phonemes possible in spoken languages. Infinite linguistic diversity seems to bring about infinite possible phonemes. But what would it mean if there were indeed infinitely many potential phonemes?

If one adopts a distinctive feature-based perspective on the phonetic targets of phonemes, then there is an *a priori* upper bound of $2^F$ contrastive sounds, where $F$ is the number of feature categories and 2 reflects the choice between presence (+) or absence (-) of that feature.^[In some distinctive feature sets, some values may be 0, i.e., undefined, underspecified, or ``for which a segment can be said 'not to care'" [@Hayes2009, pg. 91]. For example, speech sounds that are -coronal are underspecified for the features +/-anterior and +/-distributed, which capture coronal distinctions for places of articulation involving (inter-)dentals and alveolars vs palatal alveolars and retroflexes (anterior), and laminal vs apical coronals (distributed).] @Clements2009 coined this calculation *Feature Bounding*. The expected number of unique phonemes is controlled by the number of features, where additional features increase the pool of possible distinctive segments (nearly) exponentially. For example, given a feature set such as in PHOIBLE with 37 distinctive features, an upper bound on the number of segments under a strictly binary formalization is over 137 billion (2^37) feature vectors. In general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors (2^24). Crucially, this theoretical bound on phoneme diversity does not in itself preclude the possibility of infinitely many phonemes. To get there, one would only need to discover infinitely many distinctive features. But, as pointed out by @Lindblom1990a, most proponents of feature-based analysis assume a fixed, small-ish inventory of features. The question would then be: how does the feature pool grow empirically with the addition of new languages? Does it converge as assumed? Unfortunately, this question presents a host of new challenges, not the least of which concerns how to define the features themselves.

@@

Most widely accepted feature sets define features that are specified at different levels of phonetic detail: some features are more abstract (i.e., general similarities such as +/- syllabic), while others are more concrete (i.e., closer to the articulatory reality, e.g., advanced tongue root). Some are grounded in pure articulatory or acoustic phonetics, while others involve phonological properties of language. And looming in the background of these discussions, the debate of whether distinctive features are innate and universal or rather learned and emergent continues. Thus, the precise limits on what may be included as a feature are underspecified and a matter of some controversy (for more fundamental critiques of features and how they are defined, see @Lindblom1990a). Current consensus and critiques aside, there does not appear to be a principled limit to the degree of phonetic specificity that can be encoded by a feature. Making finer and finer contrasts may always produce more and more features. Thus, the exponent in the Feature Bounding calculation can itself (by hypothesis) be infinite, in line with the predictions of the models presented here, and therefore with the notion of infinite possible phonemes. That said, contemporary feature sets can capture phonological classes for place and manner of articulation, which neatly classify the phonological patterns and processes in and across languages (cf. @Mielke2008).

@@

We have so far seen that the feature space is a discrete re-coding of the continuous variability of the physical acoustic and articulatory spaces -- and so it should by way of finer and finer specification, be able to support an infinitely diverse set of phonemes. However, one objection to this account is that humans are subject to physiological limits, both in production and perception. In speech production, articulatory gestures are executed with some degree of imprecision [@Ladefoged_etal1972;@LindblomSundberg1971]. If this stochastic variability is greater than the difference between two or more phonetic targets, then no reliable distinction can be sustained (cf. @Stevens1972). In comparison, if the range of articulatory gestures is greater than their acoustic (or perceptual) reception, there is a "many-to-one" mapping between articulation and perception. In perception, the ability to discriminate acoustic stimuli is also limited both anatomically and cognitively. Differences in the signal that fall below the various thresholds for discrimination are thus ignored.

@@

<!-- In principle, Lindblom asks whether phonetic alphabets are finite (or non-finite).-->

<!-- Deriving what possible speech sounds are from an ``anthropophonic'' perspective, i.e., to derive them from the ``total sound producing potential'' of the vocal tract also allows us to investigate the evolutionary selection pressures on speech sounds [@Catford1977;@Lindblom1990a]. -->

Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond what the phonetics-phonology interface may or may not be. Thus, we believe the approaches presented here have broad applications to the study of other areas of grammar and to the cross-linguistic study of languages.

<!-- 
"Anthro"This anthropophonic perspective invites us to look for the constraints that operate in the evolutionary selection of speech sounds."

"without the anthropophonic perspective , phonetics cannot even begin to address issues such as circularity in the definition of speech sounds and the finiteness of phonetic alphabets"

we can mention the circularity and closed-class assumption of DF styles approaches. I think it is a simple elaboration if the discussion where we raise definitional issues with DF.

The anthropophonic stuff could be added as a precedent for emphasizing the closed infinity of the articulatory space. Maybe one sentence or so.

Makes sense from what you say. I guess all I could comment at the moment is that it’s striking how sound inventories are organized for (perhaps) capturing the most stable acoustic “spaces” — where the identities of sounds are recoverable irrespective of articulatory “failure”. And you seem to have an idea about what gets ignored or can be ignored in service of categorical sound distinctions.

The other point about so-called quantal vowels is that they are non-monotonic, i.e., it's very clear where they start and where they stop -- there's an abrupt perceptual distinction between them and other vowels. This is purported to be because you can articulate and [i] or [u] in myriad ways, but the acoustics are relatively stable (again because of the convergence of formants), whereas other non-quantal vowels have to be made much more precise articulatorily.

Nothing besides what I said earlier. Could mention the anthropophonemic idea (or whatever it was). Circularity of the argument for possible sound from DF approaches. Untested assumption of finite inventory of possible sounds.

Could mention the anthrop…. approach when we motivate the notion of infinite phonemes, specifically in conjunction with the infinite articulatory potential but.
-->
